---
title: "Day 17 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```

```{r}
input<-read_lines("Day17Sample.txt")
```

just pull out the coordinates for the active (#)

```{r}
startactive<-as.data.frame(matrix(ncol=3,nrow=0))
for(i in 1:length(input)){
  r<-input[i]
  for(j in 1:nchar(input[1])){
    c<-str_sub(r,j,j)
    if(c=="#"){startactive<-rbind(startactive,c(j,i,1))}}}
colnames(startactive)<-c("x","y","z")
```



## Part 1
Because this is an infinite grid - I'm going to do something stupid.

To grow -  
1 - get the places that are active (df)
2 - create a new df of all points that extend around that area by 1 in each dimension  (so a 1x1x1 will become 3x3x3)
3 - for each of the new points, find out if it will be active
4 - filter out the inactive points & repeat

Because this is starts as 8x8x1 and it only increase 6 times, it will be at most 20x20x13 which is not that big

Create a function that takes in a point and the old frame and determines whether or  
```{r}
nxtcycle<-function(cubex,cubey,cubez,olddf){
  ## is this active?
  a<-olddf%>% rowwise%>%
    filter(x==cubex,y==cubey,z==cubez)
  ### if a is 0, a is inactive
  ### if a is 1, a is active
  a<-nrow(a)
  if(a<0|a>1){cat("problem\n")}
  ##  how many of the are the neighbors active
  n<-olddf%>% rowwise%>%
    filter(x>=cubex-1,x<=cubex+1,
           y>=cubey-1,y<=cubey+1,
           z>=cubez-1,z<=cubez+1)
  n<-nrow(n)-a
  ## apply the rules
  if(a==1){
    if(n==2|n==3){nxt<-"#"}else{nxt<-"."}
  }else if(a==0){
    if(n==3){nxt<-"#"}else{nxt<-"."}}
nxt}
```

And now grow this thing:

```{r}
pocketgrow<-function(pts,reps){
 for(i in 1:reps){
   xr<-(min(pts$x)-1):(max(pts$x)+1)
   yr<-(min(pts$y)-1):(max(pts$y)+1)
   zr<-(min(pts$z)-1):(max(pts$z)+1)
   ### check all points that are 1 away from the edges.  
   newpocket<-tidyr::crossing(x=xr,y=yr,z=zr)
   newpocket<-newpocket %>% rowwise %>%
     mutate(isactive=nxtcycle(x,y,z,pts))%>%
     filter(isactive=="#")%>%
     select(-isactive)
   pts<-newpocket}
pts}
```

```{r}
part1<-pocketgrow(startactive,6)
part1<-nrow(part1)
part1
```

## Part 2

Part 2 is essentially the same, just an extra dimension

```{r}
nxtcycle4<-function(cubew,cubex,cubey,cubez,olddf){
  ## is this active?
  a<-olddf%>% rowwise%>%
    filter(x==cubex,y==cubey,z==cubez,w==cubew)
  ### if a is 0, a is inactive
  ### if a is 1, a is active
  a<-nrow(a)
  if(a<0|a>1){cat("problem\n")}
  ##  how many of the are the neighbors active
  n<-olddf%>% rowwise%>%
    filter(x>=cubex-1,x<=cubex+1,
           y>=cubey-1,y<=cubey+1,
           z>=cubez-1,z<=cubez+1,
           w>=cubew-1,w<=cubew+1)
  n<-nrow(n)-a
  if(a==1){
    if(n==2|n==3){nxt<-"#"}else{nxt<-"."}
  }else if(a==0){
    if(n==3){nxt<-"#"}else{nxt<-"."}}
nxt}
```


```{r}
pocketgrow4<-function(pts,reps){
 for(i in 1:reps){
   wr<-(min(pts$w)-1):(max(pts$w)+1)
   xr<-(min(pts$x)-1):(max(pts$x)+1)
   yr<-(min(pts$y)-1):(max(pts$y)+1)
   zr<-(min(pts$z)-1):(max(pts$z)+1)
   newpocket<-tidyr::crossing(w=wr,x=xr,y=yr,z=zr)
   newpocket<-newpocket %>% rowwise %>%
     mutate(isactive=nxtcycle4(w,x,y,z,pts))%>%
     filter(isactive=="#")%>%
     select(-isactive)
   pts<-newpocket}
pts}
```

```{r}
startactive4<-as.data.frame(matrix(ncol=4,nrow=0))
for(i in 1:length(input)){
  r<-input[i]
  for(j in 1:nchar(input[1])){
    c<-str_sub(r,j,j)
    if(c=="#"){startactive4<-rbind(startactive4,c(1,j,i,1))}}}
colnames(startactive4)<-c("w","x","y","z")
```

```{r}
part2<-pocketgrow4(startactive4,6)
part2<-nrow(part2)
part2
```
