---
title: "Day 16 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```

Start by parsing the input

```{r}
input<-read_lines("Day16Sample2.txt")
```
Split this into three parts

- the rules as a data.frame
- the my ticket (vector of numbers)
- nearby tickets (vector of strings, to start)
Rules:
```{r}
bl<-which(input=="")
### set the rules
rulesdf<-as.data.frame(matrix(nrow=0,ncol=3))
### because all of the lines in the data had the rules as two ranges
### for each line of rules, create two lines in the data frame (rule name, low, high) for each range
for(i in 1:(bl[1]-1)){
x<-input[i]
x<-unlist(str_split(x,": |-| or "))
rulesdf<-rbind(rulesdf,list(x[1],x[2],x[3]))
rulesdf<-rbind(rulesdf,list(x[1],x[4],x[5]))}
colnames(rulesdf)<-c("field","low","high")
rulesdf$high<-as.numeric(rulesdf$high)
rulesdf$low<-as.numeric(rulesdf$low)
```

My ticket
```{r}
myticket<-as.numeric(unlist(str_split(input[bl[1]+2],",")))
```

Nearby tickets
```{r}
nearbytickets<-input[(bl[2]+2):length(input)]
```

## Part 1
Collect all numbers that do not fit into any of the ranges
```{r}
invalidvalues<-function(nt,r){
  ### create vector to hold the values that don't fit
  ivals<-c()
  for(i in 1:length(nt)){
    ### for each ticket line,
    t<-as.numeric(unlist(str_split(nt[i],",")))
    for(j in 1:length(t)){
      ### check each number
      p<-t[j]
      ### check the numbers by testing them with each range (is it higher than low and lower than high)
     ok<-(sapply(1:nrow(r),function(x){
       (p>=r$low[x]&&p<=r$high[x])}))
     ### if it fits any of the ranges, it is "ok"
     ok<-any(ok)
     ### if it is not ok, add it to the list of invalid values
     if(!ok){
       ivals<-c(ivals,p)}}}
  ### return invalid values
ivals}
```

Part 1 is the sum of all invalid values
```{r}

part1<-sum(invalidvalues(nearbytickets,rulesdf))
part1
```
## Part 2

Steps (and the functions to go with them) - 

1 - actually remove the invalid lines
2 - find all possible fields for each position
3 - make the mapping unique
4 - identify the "departure" fields
5 - apply that to my ticket to get answer to part 2

First, remove the invalid lines (take in nearby tickets and rules):
```{r}
validatetickets<-function(nt,r){
  ### create a dataframe for the tickets that are "good," each field is a column
  goodtickets<-as.data.frame(matrix(nrow=0,ncol=length(unlist(str_split(nearbytickets[1],",")))))
  ## for each ticket 
  for(i in 1:length(nt)){
    ### split the ticket into pieces
    t<-as.numeric(unlist(str_split(nt[i],",")))
    ### assume it is valid
    v<-TRUE
    j<-1
    ### for each field - as long as no fields are invalid
    while(j<=length(t)&&v){
      ### as above, see if it fits into *any* of the ranges.
      p<-t[j]
      ok<-(sapply(1:nrow(r),function(x){
        (p>=r$low[x]&&p<=r$high[x])}))
      ok<-any(ok)
      ### if not - the ticket is invalid 
      if(!ok){v<-FALSE}
      j<-j+1}
    ### if the ticket  was valid, put the ticket in the frame of good tickets and move to the next ticket 
    if(v){goodtickets<-rbind(goodtickets,t)}
    i<-i+1}
  ### return a frame of good tickets
goodtickets}
```

```{r}
validtickets<-validatetickets(nearbytickets,rulesdf)
```

Using the valid tickets, find out which field each column could possibly be - takes in valid tickets and rules

```{r}
findfields<-function(rules,tickets){
  ### create the dataframe that will hold a field name and a column that might work for that field
  possibles<-as.data.frame(matrix(nrow=0,ncol=0))
  ### for each column in the list of good tickets
  for(t in 1:ncol(tickets)){
    f<-tickets[,t]
    ### because the rules are split between two lines
    for(r in seq(from=1,to=nrow(rules),by=2)){
      ### check to see which fit into the first range of the rules
      a<-f>=rules[r,2]&f<=rules[r,3]
      ### check to see which fit into the second range of the rules
      b<-f>=rules[(r+1),2]&f<=rules[(r+1),3]
      c<-a|b
      ### if all of the column fits into either of the ranges - add that as a possible mapping
      if(all(c)){possibles<-rbind(possibles,c(rules[r,1],t))}}}
  ### update the possible maps data.frame with labels on the column names
colnames(possibles)<-c("name","num")
### make the numbers numbers
possibles$num<-as.numeric(possibles$num)
possibles}
```
 
```{r}
possiblefields<-findfields(rulesdf,validtickets)
possiblefields
```

Narrow down the fields so that there's a 1-to-1 relationship

```{r}
narrowfields<-function(poss){
  ## this is sloppy
  j<-1
  ### each row is a possible field name/position number relationship
  ### as long as there are more rows than  unique position numbers, something has multiple options -
  while(nrow(poss)>length(unique(poss$num))){
    ### take possibility
    jnum<-poss$num[j]
    jname<-poss$name[j]
    ### if that column can only be one field
    if(length(which(poss$num==jnum))==1){
      ### check to see if any other columns are marked as could be that field.
      ### if so, remove those lines (because that field is now associated with the column uniquely)
      if(length(which(poss$name==jname))>1){
      poss<-poss %>% rowwise %>%
        ### mark the rows to be removed
        mutate(bye=ifelse(num!=jnum&&name==jname,F,T))%>%
        ### filter the rows
        filter(bye)%>%
        ### remove the marker
        select (-bye)}}
    ### get ready to look at a new row.
    j<-j+1
    if(j>nrow(poss)){j<-j-nrow(poss)}
  }
  ### return the 1-to-1 list of field names to ticket position numbers
  poss}
```

```{r}
fieldmap<-narrowfields(possiblefields)
fieldmap
```

Look at the fields that have "departure" in the name

```{r}
departurefields<-filter(fieldmap,str_detect(name,"departure"))
departurefields
```

apply that to my ticket and find the product of those fields

```{r}
part2<-myticket[departurefields$num]
part2<-if(length(part2)>0){prod(part2)}else{0}
part2
```

