---
title: "Day 19 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```

```{r}
#input<-read_lines("Day19Sample2.txt")
input<-read_lines("../../AoCData/AOC2020/Day19.txt")
```
Split input into rules and messages:
```{r}
bl<-which(input=="")
validrules<-input[1:(bl-1)]
recmess<-input[(bl+1):length(input)]
```




## Part 1
I thought I could just run through the rules over and over, moving them to the rule dictionary once they
are fully defined.
BUT - 0 takes too long to calculate, so stopping when I get to 8 and 11 (which are the rules that make up 0)

```{r}
rulemaker<-function(rules){
  
  rd<-dict()
  i<-1
  ##Until "0" is defined
  while(!rd$has("8")|!rd$has("11")){
    while(i>length(rules)){i<-i-length(rules)}

    ### split rule into key & value
    trk<-unlist(str_split_i(rules[i],": ",1))
    trv<-unlist(str_split_i(rules[i],": ",2))
    ### simple case - if trv is a single letter, enter it into the dictionary.
    if(str_detect(trv,"[:alpha:]")){
      rd$set(trk,str_sub(trv,2,2))
      rules<-rules[-i]
      if(i>length(rules)){i<-i-length(rules)}
      next}
    ## else
    ## check the components of the rule
    isallthere<-unlist(str_split(trv," \\| | "))
    ### if all the components in the dictionary, move forward
    isallthere<-all(sapply(isallthere,function(x){rd$has(x)}))
    ### if all the components exist, update the rule in the dictionary
    if(isallthere){
      nxttrv<-c()
      ## split along the "or"
      trv<-unlist(str_split(trv," \\| "))
      ## for each part of the "or"
      for(j in 1:length(trv)){
        tinysplit<-unlist(str_split(trv[j]," "))
        tinysplit<-lapply(tinysplit,function(x){rd$get(x)})
        nxttrv<-c(nxttrv,smushvalues(tinysplit))}
      rd$set(trk,nxttrv)
      ### remove the rule from the list of rules
      rules<-rules[-i]}
    i<-i+1
  }
  list(rd$get("8"),rd$get("11"))}
```




```{r}
smushvalues<-function(vs){
  smushed<-c()
  if(length(vs)==1){smushed<-unlist(vs)
  }else{
    a<-unlist(vs[[1]])
    for(i in 2:length(vs)){
      c<-c()
      b<-unlist(vs[[i]])
      for(j in 1:length(a)){
        for(k in 1:length(b)){
      c<-c(c,str_c(a[j],b[k]))}}
      a<-c}
      smushed<-c}
unique(smushed)}

smushvalues(list(c("a","b","g"),c("c","d"),c("e","f")))
```

```{r}
eighteleven<-rulemaker(validrules)
ruleeight<-unlist(eighteleven[[1]])
ruleeleven<-unlist(eighteleven[[2]])
```

a quick check shows that 8 always has 8 characters and 11 always has 16 characters
```{r}
min(nchar(ruleeight))
max(nchar(ruleeight))
min(nchar(ruleeleven))
max(nchar(ruleeleven))
```

So, from there, check to see which messages are valid - the valid ones will be 24 characters long, the first 8 characters in rule 8, the last 16 characters in rule 11

```{r}
part1<-sapply(recmess,function(x){
  if(nchar(x)!=24){a<-FALSE}else{a<-TRUE}
  if(!str_sub(x,1,8)%in%ruleeight){b<-FALSE}else{b<-TRUE}
  if(!str_sub(x,9,24)%in%ruleeleven){c<-FALSE}else{c<-TRUE}
  all(a,b,c)})
part1<-sum(part1)
part1
```

## Part 2
Due to the hint (thank you!), the match is
42 (n times) followed by 31 (m times) where both n & m are at least 1 *and* n is at least 1 more than m.

First - change rulemaker so it's easier to tell it when to stop

```{r}
rulemaker<-function(rules,stopit){
  rd<-dict()
  i<-1
  ##Until "0" is defined
  keepgoing<-all(sapply(stopit,function(x){rd$has(x)}))
  while(!keepgoing){
    while(i>length(rules)){i<-i-length(rules)}

    ### split rule into key & value
    trk<-unlist(str_split_i(rules[i],": ",1))
    trv<-unlist(str_split_i(rules[i],": ",2))
    ### simple case - if trv is a single letter, enter it into the dictionary.
    if(str_detect(trv,"[:alpha:]")){
      rd$set(trk,str_sub(trv,2,2))
      rules<-rules[-i]
      cat(trk,"->",str_sub(trv,2,2),"\n")
      if(i>length(rules)){i<-i-length(rules)}
      next}
    ## else
    ## check the components of the rule
    isallthere<-unlist(str_split(trv," \\| | "))
    ### if all the components in the dictionary, move forward
    isallthere<-all(sapply(isallthere,function(x){rd$has(x)}))
    ### if all the components exist, update the rule in the dictionary
    if(isallthere){
      nxttrv<-c()
      ## split along the "or"
      trv<-unlist(str_split(trv," \\| "))
      ## for each part of the "or"
      for(j in 1:length(trv)){
        tinysplit<-unlist(str_split(trv[j]," "))
        tinysplit<-lapply(tinysplit,function(x){rd$get(x)})
        nxttrv<-c(nxttrv,smushvalues(tinysplit))}
      rd$set(trk,nxttrv)
      cat(trk,"->","\n")
      ### remove the rule from the list of rules
      rules<-rules[-i]}
    i<-i+1
    keepgoing<-all(sapply(stopit,function(x){rd$has(x)}))}
  
  lapply(stopit,function(x){rd$get(x)})}
```

Then use it to get 42 & 31

```{r}
fortytwothrirtyone<-rulemaker(validrules,c("42","31"))
rule42<-unlist(fortytwothrirtyone[[1]])
rule31<-unlist(fortytwothrirtyone[[2]])
```

and as before, confirm the length of the blocks:

```{r}
min(nchar(rule42))
max(nchar(rule42))
min(nchar(rule31))
max(nchar(rule31))
```

For every message, divide it into 8 character blocks and determine if each block is in 42, 31, both, or neither.

```{r}
blocksort<-sapply(recmess,function(x){
  bs<-c()
  for(i in seq(from=1,to=nchar(x),by=8)){
    y<-str_sub(x,i,(i+7))
    if(y%in%rule42&&y%in%rule31){q<-73
    }else if(y%in%rule42){q<-42
    }else if(y%in%rule31){q<-31
    }else{q<-0}
  bs<-c(bs,q)}
bs})
blocksort[1:10]
```
a manual check says there are no "boths" which makes this easier

```{r}
isvalid<-function(v){
  checkv<-TRUE
  ### every block must match something
  if(any(is.na(v))){return(FALSE)}
  ### every first block must be 42
  if(v[1]!=42 && v[1]){return(FALSE)}
  ### every last block must be 31
  if(v[length(v)]!=31){return(FALSE)}
  threeone<-which(v==31)
  fourtwo<-which(v==42)
  ### there must be more 42s than 31s
  if(length(threeone)>=length(fourtwo)){return(FALSE)}
  s<-c()
  ### all the 42s must come before all the 31s
  for(i in 1:length(fourtwo)){s<-c(s,fourtwo[i]<threeone)}
  if(!all(s)){return(FALSE)}
checkv}

part2<-sapply(blocksort,isvalid)
for(i in 1:10){cat(recmess[i],"\t",blocksort[[i]],"\t",part2[i],"\n")}
```
```{r}
part2<-sum(part2)
part2
```

